Active Typechecking and Translation: A Safe Language-Internal Extension Mechanism

### 2013-10-12 esop14/att-esop14.tex
Up through abs reps

\section{Motivation}
When designing and implementing a new abstraction, experts typically begin by attempting to define new constructs in terms of existing language constructs.
This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a variety of interesting \emph{embedded domain-specific languages} \cite{delite}. 
Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules governing general-purpose abstractions (e.g. reasoning about units of measure requires built-in language support in F\# \cite{units-of-measure}). 
Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract. For example, regular expressions encoded using inductive datatypes are often considered overly verbose, so most functional languages support them via strings, which is less safe. 
Finally, general-purpose abstractions are also implemented in a uniform manner. Domain knowledge is thus not applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g. programmable GPUs) and distributed computing resources.

@@

Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}. This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language. At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants. This compromises both verifiability and performance. For%requires the development of an interoperability layer for every pair of DSLs. 
Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change. This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction. We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a). As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be inserted.
\begin{figure}[t]
\vspace{-15px}
\includegraphics[scale=0.415]{approaches.pdf}
\vspace{-30px}
\caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g. JVM bytecode). Users can only interface with libraries written in another language via the constructs in the common language, causing \emph{interoperability problems}. (b) With the language-internal approach, the semantics of new abstractions (i.e. the logic that governs typechecking and translation to a fixed {internal language}, here labeled \texttt{I}) can be implemented directly within so-called \emph{active libraries}. Clients can import and use these abstractions directly whenever needed.}
\vspace{-10px}
\label{approaches}
\end{figure}

@@

\emph{Internally-extensible programming languages} promise to avoid these problems by providing researchers and domain experts with a mechanism for implementing the semantics of new primitive constructs directly within libraries\todo{mention active libraries}.
%Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code. 
As a result, clients can granularly import any necessary primitive constructs when using code that relies on them, and thus achieve full safety, ease-of-use and performance due to the absence of wrappers and glue code. Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not otherwise appropriate. Libraries containing logic that is invoked at compile-time have been called \emph{active libraries} \cite{activelibraries}. We diagram this competing approach in Figure \ref{approaches}(b).

@@

												\\
\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \dint \pipe
											    \kstr \pipe
												\kunit \pipe 
												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
												\kTypeBlur \pipe \kDen \pipe 
												\kIType \pipe \kITerm
												\\
\\												
%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
%											 							&		&		&	\\
\textbf{internal terms} 				& 	\gamma	&	::=	&	\evar{x} \pipe 
												\ilam{\evar{x}}{\sigma}{\gamma} \pipe 
												\iapp{\gamma_{1}}{\gamma_{2}} \pipe
												\ifix{\evar{f}}{\sigma}{\gamma} \pipe
												\ipair{\gamma_{1}}{\gamma_{2}} \pipe 
												\ifst{\gamma} \pipe
												\isnd{\gamma}  
												\\
							& 		& 	\pipe	& 
												\iintlit \pipe \iop{\gamma_{1}}{\gamma_{2}} \pipe \iIfEq{\gamma_{1}}{\gamma_{2}}{\dint}{\gamma_{3}}{\gamma_{4}}  
												\\
 &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
%\text{deabstracted}& \gamma & ::= & \mathcal{G}[\gamma, \sigma]\\
\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe
												\dint \pipe
												\dpair{\sigma_1}{\sigma_2} \pipe
												\trepof{\tau} \pipe \dup{\tau}\\
\end{array}$$
\vspace{-10pt}
\caption{\small Syntax of \atlam. Variables $x$ are used in expressions and internal terms and are distinct from type-level variables, $\tvar{t}$. Names $\fvar{Fam}$ are family names (we assume that unique family names can be generated by some external mechanism) and $\opvar{op}$ are operator names. $\tstr{str}$ denotes string literals, $\iintlit$ denotes integer literals and $\oplus$ stands for binary operations over integers. %The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
\label{grammar}}
\end{figure}
In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise when giving users this level of control over the semantics of a language and its implementation. 
\subsection{Overview}
\todo{remove $\Theta$ from family}The grammar of \atlam~is shown in Figure \ref{grammar}. The language is a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
Types are type-level values of kind $\star$ (following System $F_{\omega}$ \cite{fomega}) and classify expressions, $e$. The type-level language also includes other kinds of terms, such as type-level functions and other type-level data structures. We include lists (required by our mechanism) as well as type-level integers, strings and products for the sake of our examples (see Sec. \ref{termination} for design constraints in general). It also includes constructs for developing extensions, which we will discuss below.
%Some kinds of type-level values (including types themselves, but excluding type-level functions) can be compared for structural equality\todo{describe this later?}.

At the top level, programs, $\rho$, consist of a series of declarations followed by an expression. Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a declaration of a new primitive type family using \textsf{family} (Sec. \ref{families}). Expressions can be either variables, lambdas, or applications of operators (Sec. \ref{operators}), and are ultimately given meaning by translation to a typed internal language. This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language. Terms of this language are denoted $\gamma$ and its types are denoted $\sigma$. Both of these productions also include additional forms containing type-level terms, $\tau$; these are used for developing extensions and during compilation and will be explained in the sections below.

\subsection{Example: G\"odel's T as an Active Type Family}
\begin{figure}[t]
\small
%\begin{lstlisting}
%family Nat of 1 ~ _.\/(int) {
%	z(i:1, a. case len a =_Z 0 then [[Nat<()> ==> 0]] else err);
%	s(i:1, a. get_last_arg a err (\a1. 
%	  dencase a1 of [[t1 ==> v1]] => 
%	    if t1 =_* Nat<()> then [[Nat<()> ==> ^(v1) + 1]] else err
%	  ow err))
%}
%\end{lstlisting}
\begin{flalign}
\label{natfam}&\family{Nat}{\kunit}{\\
\label{z}&\quad\tops{z}{\kunit}{i}{a}{
	\tapp{\tapp{\tvar{const}}{\tvar{a}}}{
		\tden{\titerm{0}}{\ttype{Nat}{\tunit}}
	}};\\
\label{s}&\quad\tops{s}{\kunit}{i}{a}{
	\tapp{
		\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x}{\kITerm}{\tlam{t}{\kTypeBlur}{
			\\
			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t}}}{\ttype{Nat}{\tunit}}}{
				\tden{\titerm{\iup{\tvar{x}}+1}}{\ttype{Nat}{\tunit}}
			}
		}}
		}
	};\\
\label{rec}&\quad\tops{rec}{\kunit}{i}{a}{
	\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x1}{\kITerm}{\tlam{t1}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x2}{\kITerm}{\tlam{t2}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x3}{\kITerm}{\tlam{t3}{\kTypeBlur}{\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
		\label{fix}&\quad\quad \tden{\titerm{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
		&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
	)}}
	}}}}}}}}}}}}
\\
&}{XXX}{i}{\titype{\dint}};\\
\label{nattype}&\pdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
\label{plus}&(\elam{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat}))}}}{
	\elam{two}{\tvar{nat}}{\\
\label{ap}	&\quad\quad \eopapp{\eopapp{plus}{two}}{two}}})
}\\
\label{recuse}&\quad (\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
	\eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
	\eop{Nat}{s}{\tunit}{r}
	}}}}})\\
\label{two}&\quad  \eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}
%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
%}
%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
	%\eop{Nat}{s}{\tunit}{r}
	%}}}}
\end{flalign}
\vspace{-15pt}
\caption{G\"odel's T in \atlam, used to calculate 2+2. The statics we implement here are shown in Appendix \ref{T}. Simple helper functions for working with argument lists ($\tvar{const}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$) are defined in Appendix \ref{helper}.}
\label{example}
\end{figure}
To make our explanation of each of the constructs in the calculus concrete, we will work through an example showing how to introduce primitive natural numbers with bounded recursion in the style of G\"odel's T \cite{pfpl}. These will be implemented internally as integers (that is, translation will always produce internal terms of internal type $\dint$). Figure \ref{example} shows how the indexed type family $\fvar{Nat}$ is defined. This family is indexed by a unit value (of kind $\kunit$), so it contains only one type, $\ttype{Nat}{\tunit}$, which we alias by defining the type-level variable $\tvar{nat}$. We define the typechecking and translation logic for the operators associated with this family (\opvar{z}, \opvar{s} and \opvar{rec}) and use these to define an add function and compute 2+2.

\subsection{Indexed Type Families and Types}\label{families}
The syntactic form $\familyDf$ declares a new primitive type family named $\fvar{Fam}$ indexed by type-level values of kind $\kappaidx$ (line \ref{natfam}). This can be compared to adding a new constructor to the compiler-internal datatype \lstinline{Type}, as suggested in Sec. \ref{evolution}. The index represents the data associated with the constructor. A base type like $\tvar{nat}$ can be thought of as being the only type in the family $\fvar{Nat}$ trivially indexed by the unit value, of kind $\kunit$, while families like $\fvar{Ntuple}$ might be indexed by a list of types, of kind $\klist{\kTypeBlur}$. A type (that is, a type-level term of kind $\kTypeBlur$) is constructed by naming a family in scope and providing a type-level term of the appropriate kind as an index. For example, $\ttype{Ntuple}{\tcons{\tvar{nat}}{\tcons{\tvar{nat}}{\tnil{\kTypeBlur}}}}$ might be the type of a pair of natural numbers. It is important that type equality be decidable, so only kinds for which equivalence coincides with syntactic equality can be used as type family indices. The main  consequence of this restriction is that indices cannot contain type-level functions.

As we will discuss further below, it is important that all expressions classified by a type compile to internal terms of consistent internal type\todo{mention compiler correctness history here?}. For this reason, we require that every type have an associated internal type. This is computed by substituting the type index for the bound variable $\tvar{i}$ in the term $\taurep$, called the \emph{representation schema}, associated with the type family and reducing to a value representing an internal type. Internal types, $\sigma$, are reified as type-level terms of kind $\kIType$ using the introductory form $\titype{\sigma}$ (as we will see, an  elimination form  is not necessary).

\subsection{Indexed Operator Families and Denotations}\label{operators}
Type families also have a collection of primitive operator families associated with them. An operator family named $\opvar{op}$ is declared using the form $\tops{op}{\kappaidx}{i}{a}{\tauop}$. Like type families, operator families are indexed by values of some specified kind, $\kappaidx$. Operators are not first-class type-level values in our calculus, so there are no restrictions related to equality. In the example in Fig. \ref{example}, all the operators are trivially indexed by the kind $\kunit$, so each family is only inhabited by one operator. However, a family like $\fvar{Ntuple}$ would have a family of projection operators, $\opvar{pr}$, indexed by a position (e.g. an integer in our calculus). A family implementing record types or object types might have a similar operator indexed by a type-level string representing the field being accessed.

An operator application is an expression of the form $\eop{Fam}{op}{\tauidx}{e_1; \ldots; e_n}$, where $n \geq 0$. The typechecking and translation of an expression of this form is controlled by the declaration of $\opvar{op}$ in family $\fvar{Fam}$, as determined by the type-level term $\tauop$. It must evaluate to a \emph{denotation}, which is a type-level value of kind $\kDen$, when the index, $\tauidx$, and a list containing the denotations recursively assigned to each argument, $e_1$ through $e_n$, are bound to $\tvar{i}$ and $\tvar{a}$, respectively.

There are two forms of denotations. A \emph{valid denotation} has the form $\tden{\tau_1}{\tau_2}$, where $\tau_1$ represents the \emph{translation} of an expression to an internal term and $\tau_2$ is the type. The translation must have kind $\kITerm$ of  reified internal language terms. The only introductory form for this is $\titerm{\gamma}$, just as with $\kIType$. There is no elimination form. Types, as before, have kind $\kTypeBlur$.
%(indeed, in \atlam, terms are never  syntactically analyzed directly by extensions, unlike macro systems and other forms of term rewriting; see Sec. \ref{related-work})
If a type cannot be assigned to an operation according to the specification of the operator being implemented, then the \emph{error denotation}, $\terr$, should be returned instead of a valid denotation. In a practical implementation, a specialized error message and other diagnostic information could be associated with $\terr$, but we omit such details for the sake of simplicity in our core calculus. Terms of kind $\kDen$ can be \textsf{case} analyzed to determine if they are valid, and if so, to extract the translation and type.

@@

\inferrule[att-exp]{
	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\ghat}\\
	\eraseX{\ghat}{\gamma}
}{
	\pcompilesX{e}
}
\end{mathpar}
$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\ghat}}}$
~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
\begin{mathpar}
\inferrule[att-var]{ }{
	\ecompiles{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
}

>>>>

\section{Motivation}
When designing and implementing a new abstraction, experts typically begin by attempting to define new constructs in terms of existing language constructs.
This approach is often effective because modern {general-purpose} abstraction mechanisms, like inductive datatypes and object systems, are highly expressive.  
For example, the Delite framework leverages Scala's powerful general-purpose mechanisms to enable a variety of interesting \emph{embedded domain-specific languages} \cite{delite}. 
Unfortunately, there remain some situations of interest where general-purpose abstractions fall short. 
For instance, it is difficult to adequately encode advanced type systems in terms of the simpler rules governing general-purpose abstractions (e.g. reasoning about units of measure requires built-in language support in F\# \cite{units-of-measure}). 
Even if a full encoding is possible, it may not be useful if it is overly verbose or unnatural, or if the error messages are overly abstract. For example, regular expressions encoded using inductive datatypes are often considered overly verbose, so most functional languages support them via strings, which is less safe. 
Finally, general-purpose abstractions are implemented in a uniform manner. Domain knowledge is not easily applied to eliminate overhead or perform optimizations, and implementations designed for typical application workloads may not be satisfactory in parts of a program where performance is a key criteria, particularly when targeting heterogeneous hardware platforms (e.g. programmable GPUs) and distributed computing resources. Using variants of familiar abstractions backed by specialized implementation strategies are quite useful in these cases.

@@

Unfortunately, this leads to a critical problem at language boundaries: a library's external interface must only use constructs that can reasonably be expressed in \emph{all possible client languages}. This discourages languages from including constructs that rely on statically-checked invariants stronger than those supported by their underlying implementation in the common intermediate language. At best, constructs like these can be exposed by generating a wrapper where run-time checks have been inserted to guarantee these invariants. This compromises both verifiability and performance. For%requires the development of an interoperability layer for every pair of DSLs. 
Moreover, this approach exposes the internals of an implementation to clients, making the abstraction awkward to work with and causing code breakage when implementation details change. This defeats a primary purpose of high-level programming languages: hiding low-level details from clients of an abstraction. We diagram this fundamental \emph{interoperability problem} in Figure \ref{approaches}(a). As an example, F\#'s type system prevents \lstinline{null} values from occurring within data structures, but because it's type system is not available when calling into F\# code from another language, like C\#, run-time null checks must still be included in the implementation.
\begin{figure}[t]
\vspace{-15px}
\includegraphics[scale=0.415]{approaches.pdf}
\vspace{-30px}
\caption{(a) With the language-oriented approach, different primitive abstractions are packaged into separate languages that extend and target a common intermediate language (e.g. JVM bytecode). Users can only interface with libraries written in another language via the constructs in the common language, causing \emph{interoperability problems}. (b) With the language-internal approach, the semantics of new abstractions (i.e. the logic that governs typechecking and translation to a fixed {internal language}, here labeled \texttt{I}) can be implemented directly within so-called \emph{active libraries}. Clients can import and use these abstractions directly whenever needed.}
\vspace{-10px}
\label{approaches}
\end{figure}

@@

\emph{Internally-extensible programming languages} promise to avoid these problems by providing researchers and domain experts with a mechanism for implementing the semantics of new primitive constructs directly within libraries.
%Developers of libraries need only determine whether they are appropriate for their domain, without also considering whether these constructs can be exposed in terms of abstractions appropriate to client code. 
As a result, clients can granularly import any necessary primitive constructs when using code that relies on them, and thus achieve full safety, ease-of-use and performance due to the absence of wrappers and glue code. Providers of components thus need only consider whether primitives that they use are appropriate for their domain, without also considering whether their code might be used in a context where these primitives are not otherwise appropriate. Libraries containing logic that is invoked at compile-time, as extension logic would be, have been called \emph{active libraries} \cite{activelibraries}. We adopt this terminology and diagram this competing approach in Figure \ref{approaches}(b).

@@

												\\
\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\kappa_1}{\kappa_2} \pipe \klist{\kappa} \pipe \dint \pipe
											    \kstr \pipe
												\kunit \pipe 
												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
												\kTypeBlur \pipe \kDen \pipe 
												\kIType \pipe \kITerm
												\\
\\												
%\textbf{ops signature}			& \Theta	&	::=	&	\kOpEmpty \pipe \kOp{\Theta}{op}{\kappai}\\
%											 							&		&		&	\\
\textbf{internal terms} 				& 	\gamma	&	::=	&	\evar{x} \pipe 
												\ilam{\evar{x}}{\sigma}{\gamma} \pipe 
												\iapp{\gamma_{1}}{\gamma_{2}} \pipe
												\ifix{\evar{f}}{\sigma}{\gamma} \pipe
												\ipair{\gamma_{1}}{\gamma_{2}} \pipe 
												\ifst{\gamma} \pipe
												\isnd{\gamma}  
												\\
							& 		& 	\pipe	& 
												\iintlit \pipe \iop{\gamma_{1}}{\gamma_{2}} \pipe \iIfEq{\gamma_{1}}{\gamma_{2}}{\dint}{\gamma_{3}}{\gamma_{4}}  
												\\
 &  & \pipe & \tvalof{\tau_1}{\tau_2} \pipe \iup{\tau} \\
%\text{deabstracted}& \gamma & ::= & \mathcal{G}[\gamma, \sigma]\\
\textbf{internal types}			&	\sigma	&	::=	&    \darrow{\sigma_1}{\sigma_2} \pipe
												\dint \pipe
												\dpair{\sigma_1}{\sigma_2} \pipe
												\trepof{\tau} \pipe \dup{\tau}\\
\end{array}$$
\vspace{-10pt}
\caption{\small Syntax of \atlam. Variables $x$ are used in expressions and internal terms and are distinct from type-level variables, $\tvar{t}$. Names $\fvar{Fam}$ are family names (we assume that unique family names can be generated by some external mechanism) and $\opvar{op}$ are operator names. $\tstr{str}$ denotes string literals, $\iintlit$ denotes integer literals and $\oplus$ stands for binary operations over integers. %The productions related to the internal language are written using generators $\mathcal{G}$ and $\mathcal{S}$ to avoid duplicating the syntax of common terms.
\label{grammar}}
\end{figure}
In this section, we will develop a core calculus, called @$\lambda$ for the ``actively-typed lambda calculus'', by way of a semantics and a simple example, and discuss how it addresses the safety concerns that arise when giving users this level of control over the semantics of a language and its implementation. 
\subsection{Overview}
The grammar of \atlam~is shown in Figure \ref{grammar}. The language is a simply-typed lambda calculus with simply-kinded type-level computation. Kinds, $\kappa$, classify type-level terms, $\tau$. 
Types are type-level values of kind $\star$ (following System $F_{\omega}$ \cite{fomega}) and classify expressions, $e$. The type-level language also includes other kinds of terms, such as type-level functions, lists (required by our mechanism) and integers, strings and products for the sake of our examples (see Sec. \ref{termination} for a discussion on other acceptable kinds of type-level data). It also includes constructs for developing extensions -- denotations and reified internal terms and types -- which we will discuss in the sections below.

At the top level, programs, $\rho$, consist of a series of declarations followed by an expression. Declarations can be either bindings of type-level terms to type-level variables using \textsf{def} or a declaration of a new primitive type family using \textsf{family}. Expressions can be either variables, lambdas, or applications of operators, and are ultimately given meaning by translation to a typed internal language. This language has been chosen, for simplicity, to be a variant of Plotkin's PCF with primitive integers and products, but in practice would include other constructs consistent with its role as a high-level intermediate language. The grammars of internal terms, $\gamma$, and internal types, $\sigma$, also include special forms containing type-level terms; these are used for developing extensions and during compilation and will be erased before compilation ends, as we will explain below.

\subsection{Example: G\"odel's T as an Active Type Family}
\begin{figure}[t]
\small
%\begin{lstlisting}
%family Nat of 1 ~ _.\/(int) {
%	z(i:1, a. case len a =_Z 0 then [[Nat<()> ==> 0]] else err);
%	s(i:1, a. get_last_arg a err (\a1. 
%	  dencase a1 of [[t1 ==> v1]] => 
%	    if t1 =_* Nat<()> then [[Nat<()> ==> ^(v1) + 1]] else err
%	  ow err))
%}
%\end{lstlisting}
\begin{flalign}
\label{natfam}&\family{Nat}{\kunit}{\\
\label{z}&\quad\tops{z}{\kunit}{i}{a}{
	\tapp{\tapp{\tvar{const}}{\tvar{a}}}{
		\tden{\titerm{0}}{\ttype{Nat}{\tunit}}
	}};\\
\label{s}&\quad\tops{s}{\kunit}{i}{a}{
	\tapp{
		\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x}{\kITerm}{\tlam{t}{\kTypeBlur}{
			\\
			&\quad\quad\tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t}}}{\ttype{Nat}{\tunit}}}{
				\tden{\titerm{\iup{\tvar{x}}+1}}{\ttype{Nat}{\tunit}}
			}
		}}
		}
	};\\
\label{rec}&\quad\tops{rec}{\kunit}{i}{a}{
	\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x1}{\kITerm}{\tlam{t1}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
	&\quad\quad\tapp{\tapp{\tvar{pop}}{\tvar{a}}}{\tlam{x2}{\kITerm}{\tlam{t2}{\kTypeBlur}{\tlam{a}{\klist{\kDen}}{\\
	&\quad\quad \tapp{\tapp{\tvar{pop\_final}}{\tvar{a}}}{\tlam{x3}{\kITerm}{\tlam{t3}{\kTypeBlur}{\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t1}}}{\ttype{Nat}{\tunit}}}{(\\
	&\quad\quad \tapp{\tapp{\tapp{\tvar{check\_type}}{\tvar{t3}}}{\ttype{Arrow}{(\ttype{Nat}{\tunit},\ttype{Arrow}{(\tvar{t2},\tvar{t2})})}}}{\\
		\label{fix}&\quad\quad \tden{\titerm{\iapp{(\ifix{f}{\darrow{\dint}{\trepof{\tvar{t2}}}}{\ilam{x}{\dint}{\\
		\label{lastop}&\quad\quad\quad \iIfEq{x}{0}{\dint}{\iup{\tvar{x2}}}{\iapp{\iapp{\iup{\tvar{x3}}}{(x-1)}}{(\iapp{f}{(x-1)})}}}})}{\iup{\tvar{x1}}}}}{\tvar{t2}}
	)}}
	}}}}}}}}}}}}
\\
&}{XXX}{i}{\titype{\dint}};\\
\label{nattype}&\pdef{nat}{\kTypeBlur}{\ttype{Nat}{\tunit}}{\\
\label{plus}&(\elam{plus}{\ttype{Arrow}{(\tvar{nat}, \ttype{Arrow}{(\tvar{nat}, \tvar{nat}))}}}{
	\elam{two}{\tvar{nat}}{\\
\label{ap}	&\quad\quad \eopapp{\eopapp{plus}{two}}{two}}})
}\\
\label{add}&\quad (\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{
	\eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
	\eop{Nat}{s}{\tunit}{r}
	}}}}})\\
\label{two}&\quad  \eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}
%{\\&\elet{two}{\tvar{nat}}{\eop{Nat}{s}{\tunit}{\eop{Nat}{s}{\tunit}{\eop{Nat}{z}{\tunit}{ }}}}{\eapp{\eapp{plus}{two}}{two}}}
%}
%\elam{x}{\tvar{nat}}{\elam{y}{\tvar{nat}}{\\
	%&\quad \eop{Nat}{rec}{\tunit}{x; y; \elam{p}{\tvar{nat}}{\elam{r}{\tvar{nat}}{
	%\eop{Nat}{s}{\tunit}{r}
	%}}}}
\end{flalign}
\vspace{-15pt}
\caption{G\"odel's T in \atlam, used to calculate 2+2. The statics we implement here are shown in Appendix \ref{T}. Simple helper functions for working with argument lists ($\tvar{const}$, $\tvar{pop}$, $\tvar{pop\_final}$) and types ($\tvar{check\_type}$) are defined in Appendix \ref{helper}.}
\label{example}
\vspace{-10pt}
\end{figure}
To make our explanation of each of the constructs in the calculus concrete, we will work through an example showing how to introduce primitive natural numbers with bounded recursion in the style of G\"odel's T \cite{pfpl}. These will be implemented internally as integers (internal terms of internal type $\dint$). Figure \ref{example} shows how the indexed type family $\fvar{Nat}$ is defined. This family is indexed by a unit value (of kind $\kunit$), so it contains only one type, $\ttype{Nat}{\tunit}$, which we alias on line \ref{nattype} by defining the type-level variable $\tvar{nat}$. We define the typechecking and translation logic for the operators associated with this family (\opvar{z}, \opvar{s} and \opvar{rec}) on lines \ref{z}-\ref{lastop} and use these to define a $plus$ function on line \ref{add} and compute 2+2.

\subsection{Indexed Type Families and Types}\label{families}
The syntactic form $\familyDf$ declares a new primitive type family named $\fvar{Fam}$ indexed by type-level values of kind $\kappaidx$ with representation schema $\tvar{i}.\taurep$ and operators $\theta$. This can be compared to adding a new constructor to the compiler-internal datatype \lstinline{Type}, as suggested in Sec. \ref{evolution}. 
The index represents the data associated with the constructor. A base type like $\tvar{nat}$ can be thought of as being the only type in the family $\fvar{Nat}$ trivially indexed by the unit value, of kind $\kunit$, while families like $\fvar{Ntuple}$ might be indexed by a list of types, having kind $\klist{\kTypeBlur}$. 
A type (that is, a type-level term of kind $\kTypeBlur$) is constructed by naming a family in scope and providing a type-level term of the appropriate kind as an index. For example, $\ttype{Ntuple}{\tcons{\tvar{nat}}{\tcons{\tvar{nat}}{\tnil{\kTypeBlur}}}}$ might be the type of a pair of natural numbers. It is important that type equality be decidable, so only kinds for which equivalence coincides with syntactic equality can be used as type family indices. The main  consequence of this restriction is that indices cannot contain type-level functions. Given a type, its family can be \textsf{case} analyzed to extract its index.

\subsection{Representation Schemas}
As we will discuss further below, it is important that all expressions classified by a type compile to consistently-typed internal terms\todo{mention compiler correctness history here?}. For this reason, we require that every type have associated with it a single internal type. This is computed by substituting the type index for the bound variable $\tvar{i}$ in the term $\taurep$, called the \emph{representation schema}, associated with the type family and evaluating to a value representing an internal type. Internal types, $\sigma$, are reified into type-level terms of kind $\kIType$ using the introductory form $\titype{\sigma}$.

\subsection{Indexed Operator Families and Denotations}\label{operators}
Type families also have a collection of primitive operator families associated with them. An operator family named $\opvar{op}$ is declared using the form $\tops{op}{\kappaidx}{i}{a}{\tauop}$. Like type families, operator families are indexed by values of some kind, $\kappaidx$. Operators are not first-class type-level values in our calculus, so there are no restrictions on this kind related to equality. In the example in Fig. \ref{example}, all the operators are trivially indexed by the kind $\kunit$, so each family only contains one operator. However, a type family like $\fvar{Ntuple}$ would be equipped with a family of projection operators, $\opvar{pr}$, indexed by a position (e.g. an integer in our calculus). A family implementing record types or object types might have a similar operator indexed by a type-level string representing the field being accessed.

To select an operator from a family by providing an index and apply it to $n$ arguments, the grammar provides a uniform form: $\eop{Fam}{op}{\tauidx}{e_1; \ldots; e_n}$, where $n \geq 0$. The typechecking and translation of an expression of this form is controlled by the implementation, $\tauop$. It must evaluate to a \emph{denotation}, which is a type-level value of kind $\kDen$, when given the operator index, $\tauidx$, and a list constructed from the denotations recursively assigned to each argument, $e_1$ through $e_n$. 
There are two forms of denotations that an expression can have. A \emph{valid denotation} has the form $\tden{\tauiterm}{\tautype}$, where $\tauiterm$ represents the \emph{translation} of the expression to an internal term and $\tautype$ is the type assigned to it. Internal terms are reified as type-level terms using the form $\titerm{\gamma}$, and have kind $\kITerm$ (similar to $\kIType$). It is important to note that there are no elimination form for reified terms or types.
%(indeed, in \atlam, terms are never  syntactically analyzed directly by extensions, unlike macro systems and other forms of term rewriting; see Sec. \ref{related-work})
If a type cannot be assigned to an operation according to the specification of the operator being implemented, then the \emph{error denotation}, $\terr$, is returned instead of a valid denotation. In a more practical implementation, a specialized error message and other diagnostic information could be associated with $\terr$, but we omit such details for the sake of simplicity in our core calculus. Terms of kind $\kDen$ can be \textsf{case} analyzed to determine if they are valid or errors, and if valid, to extract the translation and type.

@@

\inferrule[att-exp]{
	\ecompiles{\emptyctx}{\fvalCtx}{e}{\tau}{\gabs}\\
	\eraseX{\gabs}{\gamma}
}{
	\pcompilesX{e}
}
\end{mathpar}
$\fbox{\inferrule{}{\ecompilesX{e}{\tau}{\gabs}}}$
~~~$\etCtx ::= \emptyctx \pipe \etCtxX{x}{\tau}$
\begin{mathpar}
\inferrule[att-var]{ }{
	\ecompiles{\etCtxX{x}{\tau}}{\fvalCtx}{\evar{x}}{\tau}{\evar{x}}
}
